The Design and Implementation of the Synchronous Language \CEU

### 2016-09-26 paper.tex
(>) timers

Activities that involve reactions to \emph{wall-clock time}%
\footnote{
By wall-clock time we mean the passage of time from the real world, measured in 
hours, minutes, etc.
}
appear in typical patterns of embedded development, such as timeout watchdogs 
and sensor samplings.
%
However, the interaction between system clocks and programs is not absolutely
precise, a fact that is usually ignored in the development process.
We define the difference between a requested timeout and the actual expiring 
time as the \emph{residual delta time (delta)}.
Without explicit manipulation, the recurrent use of timed activities in a row 
(or in a loop) may accumulate a considerable amount of deltas that can lead to 
incorrect behavior in programs.

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
In the example in Figure~\ref{lst.timers}.a, suppose that after the first 
\code{await} request, the underlying system gets busy and takes 15ms to notify 
\CEU.
The scheduler will notice that the \code{await 10ms} (ln. 2) has not only 
already expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1} (ln. 3), and invokes 
\code{await 1ms} (ln. 4).
As the current delta is still higher than the requested timeout (i.e. $5ms > 
1ms$), the trail is rescheduled for execution, now with \code{delta=4ms}.

@@

\CEU also considers that time is a physical quantity that can be added and 
compared.
For instance, in the example in Figure~\ref{lst.timers}.b, although the 
scheduler cannot guarantee that the first trail terminates exactly in 11ms (ln.  
2,4), it can at least ensure that the program always terminates with 
\code{v=1}.
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail (ln. 2--5) is guaranteed to terminate before 
the second trail (ln. 7--8), because $10+1 < 12$.
%
A similar program in a language without first-class support for timers would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU also performs concurrency checks to detect trails that, when reordered,
change the observable behavior of the program, i.e., trails that actually rely
on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the concurrency checks of \CEU.
%
\CEU also guarantees deterministic behavior for timer compositions by adjusting 
for inaccuracies in the system clock.

>>>>

Activities that involve reactions to \emph{wall-clock time}%
\footnote{
By wall-clock time we mean the passage of time from the real world, measured in 
hours, minutes, etc.
}
appear in typical patterns of embedded development~\cite{esterel.delays}, such
as timeout watchdogs and sensor samplings.
%
However, the interaction between system clocks and programs is not absolutely
precise, a fact that is usually ignored in the development process.
We define the difference between a requested timeout and the actual expiring 
time as the \emph{residual delta time (delta)}.
Without explicit manipulation, the recurrent use of timed activities in a row 
(or in a loop) may accumulate a considerable amount of deltas that can lead to 
incorrect behavior in programs.

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
Due to platform overheads and occasional slow reactions, an \code{await 10ms}
suspends a line of execution for \emph{at least} 10 milliseconds.
In the example in Figure~\ref{lst.timers}.a, suppose that after the first 
\code{await} request, the underlying system gets busy and takes 15ms to notify 
\CEU.
The scheduler will notice that the \code{await 10ms} (ln. 2) has not only 
already expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1} (ln. 3), and invokes 
\code{await 1ms} (ln. 4).
As the current delta is still higher than the requested timeout (i.e. $5ms > 
1ms$), the trail is rescheduled for execution, now with \code{delta=4ms}.

@@

Delta compensation also implies that timers can be added and compared when
reasoning about time as a physical quantity.
For instance, in the example in Figure~\ref{lst.timers}.b, although the 
scheduler cannot guarantee that the first trail terminates exactly in 11ms (ln.  
2,4), it can at least ensure that the program always terminates with 
\code{v=1}.
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail (ln. 2--5) is guaranteed to awake first and
terminate the \code{par/or} before the second trail executes (ln. 7--8),
because $10+1 < 12$.
%
A similar program in a language without first-class support for timers would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU also performs concurrency checks to detect trails that, when reordered,
change the observable behavior of the program, i.e., trails that actually rely
on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the concurrency checks of \CEU.

\emph{Esterel+Delay}~\cite{esterel.delays} is a non-intrusive extension to
Esterel to program in terms of physical time with \code{delay} statements.
The global transformation relies on a \emph{platform statement} the specifies
the available timers in the system (e.g., sample-driven or interrupt-driven).
It then expands \code{delay} statements into existing Esterel statements in a
way that the desired semantics can be realised in the platform.
%
\CEU makes physical time a special input event that feeds the runtime with an
associated time to elapse which is decremented from all awaiting trails.
The compensation scheme of \CEU guarantees that trails awake in the correct
order and that errors are not propagated on subsequent awaits.
Interrupt-driven timers are also supported with a hook callback that the
runtime calls whenever the program awaits an earlier timer.
